Hierarchy Builder: algebraic~hierarchies made easy in Coq with Elpi

### 2020-04-24 hb.tex
splitting, refactoring and clarifications in Sect 3.1

\definecolor{dkgreen}{rgb}{0, 0.502, 0}

@@

\ccsdesc[100]{General and reference~General literature}
\ccsdesc[100]{General and reference}%TODO mandatory: Please choose ACM 2012 classifications from https://dl.acm.org/ccs/ccs_flat.cfm
\supplement{Source code of the Coq package: \url{https://github.com/math-comp/hierarchy-builder}}
\keywords{Algebraic Hierarchy, Packed Classes, Coq, Elpi, Metaprogramming, $\lambda$Prolog}%TODO mandatory; please add comma-separated list of keywords

@@

\relatedversion{}%optional, e.g. full version hosted on arXiv, HAL, or other respository/website
%\relatedversion{A full version of the paper is available at \url{...}.}

\supplement{}%optional, e.g. related research data, source code, ... hosted on a repository like zenodo, figshare, GitHub, ...

@@

%Editor-only macros:: begin (do not touch as author)%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\EventEditors{John Q. Open and Joan R. Access}
\EventNoEds{2}
\EventLongTitle{42nd Conference on Very Important Topics (CVIT 2016)}
\EventShortTitle{CVIT 2016}
\EventAcronym{CVIT}
\EventYear{2016}
\EventDate{December 24--27, 2016}
\EventLocation{Little Whinging, United Kingdom}
\EventLogo{}
\SeriesVolume{42}
\ArticleNo{23}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

@@

@@

\begin{definition}[\M{}, \mixins{}]
A \emph{\mixin{}} \(m \in \M{}\) is a Coq record with one or more parameters.
The first parameter must be a \coq{(T : Type)}, while the other parameters
are \mixins{} applied to \coq{T} and possibly $n$ other distinct \mixins{}. I.e.
\begin{coqcode}
Record #$m$# (T : Type) #\(\pmp{n}\)# : Type := { .. }.
\end{coqcode}
where \(\pmp{n} = \vect{\left(p_i : m_i\ T\ p_{\sigma(i,1)}\ \ldots\ p_{\sigma(i,n_i)}\right)}_{i \in \enum1n}\).
For all \(i \in \enum1n\) we have \(m_i \in \M{}\) and its
arguments consist in $n_i$ of the previously quantified
mixin parameters $p_k$, i.e. for all \(k \in \enum1{n_i}\), we
have \(\sigma(i,k) \in \enum1{i-1}\).
\end{definition}

@@

\begin{remark}For all \(i \in \enum1n\), we have
\(\dep{}(m_i) = \enum{m_{\sigma(i,1)}}{m_{\sigma(i,m_i)}}\).
\end{remark}

@@

\begin{proposition}
\dep{} is transitively closed since records are well typed in the
empty context and describes a DAG since Coq does not admit circular
definitions: \(\dep{}^\star(\dep{}^\star(M)) \subseteq \dep{}^\star(M)\)
\end{proposition}

\begin{definition}[\factories{} \F{}, \mixinbuilders{} \B{} and \from{}]
  A \emph{\mixinbuilder{}} \(\mu \in \B{}\) is any function which parameters are the carrier
  type \coq{(T: Type)}, an arbitrary number of distinct \mixins{}
  \(\enum{m_1}{m_n}\) and an additional argument \(f\), which we call
  a \emph{\factory{}}, and which depends on the \mixins{}. Its return type is
  a \mixin{} \(m_{n+1} \in \M{}\) i.e.
\begin{coqcode}
Definition f (T : Type) #$\pmp{n}$# : Type := ...
Definition #$\mu$# (T : Type) #$\pmp{n}$#: #\(f\ T\ p_1 \ldots p_n\)# -> #\(m_{n+1}\ T\ p_{\sigma(n+1,1)}\ \ldots\ p_{\sigma(n+1,n_i)}\)# := ...
\end{coqcode}

  We define \(\from{} (f, m_{n+1}) = \mu\) to be the (unique) \mixinbuilder{}
  for \(m_{n+1}\) from the \factory{} \(f\).
\end{definition}

\begin{definition}[$\provides{}$, $\requires{}$]
  We define the following functions \begin{itemize}
  \item\( \requires{}(f) = \enum{m_1}{m_n}  \in \powerset\M{}\)
    is the set of mixins a \factory{} \(f \in \F{}\) depends on.
    As a consequence all the
    \mixinbuilders{} of a given \factory{} have the same set of
    dependencies.
  \item\( \provides{}(f) = \{m | \from{}(f,m) \textrm{ is defined}\}
    \in \powerset\M{}\) the set of mixins that a factory
    \(f \in \F{}\) provides, through its \mixinbuilders{}.
  \end{itemize}
Moreover the following properties hold for all \(f \in \F{}\):
  %\begin{proposition}
  %  They validate the following properties for all \(f \in \F{}\):
    \begin{enumerate}
      \item\( \dep{}^\star(\requires{}(f)) \subseteq \requires{}(f)\)
      \item\( \requires{}(f) \cap \provides{}(f) = \emptyset \)
      \item\( \forall m \in \provides{}(f),~ \from{}(f,m) \) is defined
    \end{enumerate}
  %\end{proposition}
\end{definition}

@@

\begin{proposition}[\( \M{} \subseteq \F{} \)] For all \(m \in \M{}\)
  we have \(\requires{}(m) = \dep{}(m)\), \(\provides{}(m) = \{m\}\)
  and
  \(\from{}(m,m) = \left(\coq{fun T }\ \pmp{n} \ \coq{(x :}\ m\
  \coq{T }\ p_1\ \ldots p_n\ \coq{) => x}\right) \in \B{}.\)
\end{proposition}

\begin{command} to declare a new mixin:
\begin{coqcode}
#\hbmixin{}# Record M T of #$f_1\ \ldots\ f_n$# := { .. }.
\end{coqcode}
  Creates a module \coq{M} with a record
  \coq{M}, which depends on the \mixins{}
  \(\requires{}(\coq{M}) =
  \provides{}^\star\left(\enum{f_1}{f_n}\right)\), and registers this
  new record both as a \mixin{} and a \factory{}. It also creates a
  \phantterm{} \coq{M.Build.}
\end{command}

\begin{command} to declare an instance: \hbinstance{}
  \(X\ b_1 \ldots b_k\), synthesizes terms corresponding to all the
  \mixins{} that can be built from the $b_i$. Indeed if
  \(b_i : f_i\ T\ \ldots\), then this command creates elements of type
  \(\provides{}^\star\left(\enum{f_1}{f_k}\right).\) This command also
  generates unification hints as described in
  Section~\ref{sec:target-lang}.
\end{command}

\begin{command} to declare a new \factory{} and generate new \mixinbuilders:
\begin{coqcode}
#\hbfactory{}# Record M T of #$f_1\ \ldots\ f_n$# := { .. }.
#\hbbuilders{}# Context T (a : M A).

  Definition #$b_{n+1}$# : #$f_{n+1}$# T .. := ...
  #\hbinstance{}# T #$b_{n+1}$#.
  ..
  Definition #$b_{n+k}$# : #$f_{n+k}$# T .. := ...
  #\hbinstance{}# T #$b_{n+k}$#.
#\hbend{}#.
\end{coqcode}
  Creates a module \coq{M} with a record
  \coq{M}, which depends on the mixins
  \(\requires{}(\coq{M}) =
  \provides{}^\star\left(\enum{f_1}{f_n}\right)\) and registers this
  new record as a factory. We then use the \factoryinstances{}
  \(b_{n+1}, \ldots, b_{n+k}\) provided by the user and derive
  \mixinbuilders{} from them, so that
  \(\provides{}(\coq{M}) = \provides{}^\star\left(\enum{f_{n+1}}{f_{n+k}}\right).\)

  It is thus necessary that
  \(\requires{}^\star\left(\enum{f_{n+1}}{f_{n+k}}\right) \subseteq
  \provides{}^\star\left(\enum{f_1}{f_n}\right).\)
\end{command}

Note that the \(b_i\) are not \mixinbuilders{} since their return types
are not necessarily \mixins{}, but could be a \factory{}. However, since all
\factories{} provide \mixin{} through \mixinbuilders{}, we can
obtain \mixinbuilders{} out of each \(b_i\) by function composition.

@@

\begin{definition}[\C{}, class]
A \class{} \coq{c} \(\in \C{}\) is a Coq record with one parameter
\coq{(T : Type)}. The type of each field is a \mixin{} in \M{} applied to
\coq{T} and, if needed, any number of other fields:
\begin{coqcode}
Record #$c$# (T : Type) := { #\(p_1 : m_1\ T\)#; .. ; #\(p_i : m_i\ T\ p_{\sigma(i,1)}\ \ldots p_{\sigma(i,n_i)}\)#; ..}.
\end{coqcode}
\end{definition}

@@

\begin{definition}[\Str{} Structure]
A \structure{} \coq{s} \(\in \Str{}\) is a dependent pair: a Coq record where
the value of the first field occurs in the type of the second.
The first field is \coq{(sort : Type)} and the second field is
\coq{(class : c sort)} for some \coq{c} \(\in \C{}\).
As a consequence \structures{} are in bijection with \classes{}.
\end{definition}

\begin{command} to declare a class and structure:
  \hbstructure{} \coq#Definition M := { A of# \(f_1 \ldots f_n\) \coq#}.#
  crafts a class \coq{M.class_of} $\in \C{}$ where
  $\cdef{}(c) = \provides{}^\star\left(\enum{f_1}{f_n}\right) $ and the
  corresponding structure \coq{M.type} $\in \Str{}$, together with
  unification hints as described in Section~\ref{sec:target-lang}.
\end{command}

>>>>

\definecolor{dkgreen}{rgb}{0, 0, 1}

@@

\ccsdesc{Software and its engineering~Formal language definitions}
\ccsdesc{Theory of computation~Type theory}
\ccsdesc{Theory of computation~Constraint and logic programming}
\ccsdesc{Computing methodologies~Symbolic and algebraic manipulation}

\supplement{Source code of the Coq package: \url{https://github.com/math-comp/hierarchy-builder}}
\keywords{Algebraic Hierarchy, Packed Classes, Coq, Elpi, Metaprogramming, $\lambda$Prolog}%TODO mandatory; please add comma-separated list of keywords

@@

\relatedversion{System Description}%optional, e.g. full version hosted on arXiv, HAL, or other respository/website
%\relatedversion{A full version of the paper is available at \url{...}.}

@@

%Editor-only macros:: begin (do not touch as author)%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\EventEditors{Zena M. Ariola}
\EventNoEds{1}
\EventLongTitle{5th International Conference on Formal Structures for Computation and Deduction (FSCD 2020)}
\EventShortTitle{FSCD 2020}
\EventAcronym{FSCD}
\EventYear{2020}
\EventDate{June 29--July 5, 2020}
\EventLocation{Paris, France}
\EventLogo{}
\SeriesVolume{167}
\ArticleNo{8}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

@@

\newtheoremstyle{commands}{\topsep}{\topsep}{}{0pt}{\sffamily\bfseries}{ }{5pt plus 1pt minus 1pt}%
{$\blacktriangleright$ \thmname{#1} {\coq{#3}}}
\theoremstyle{commands}
\newtheorem*{commands}{Coq commands}

@@

In this section, we call ``distinguished'' a Coq definition or record
that the developer of a library has labeled ``\mixin{}'', ``\factory{}'',
``\mixinbuilder{}'', ``\class{}'' or ``\structure{}''.

\begin{definition}[\M{}, \mixins{}]
  A \emph{\mixin{}} \(m \in \M{}\) is a distinguished Coq record with one or more
  parameters.  The first parameter must be a \coq{(T : Type)}, while
  the other parameters are \mixins{} \(\left(m_i\right)_{i \in \enum1n}\), each of which
  is applied to \coq{T} and possibly previous mixin variables. I.e.
\begin{coqcode}
Record #$m$# (T : Type) #\(\pmp{n}\)# : Type := { .. }.
\end{coqcode}
where \(\pmp{n} = (p_1 : m_1\ T) \ldots \left(p_n : m_n\ T\ p_{\sigma(n,1)}\ \ldots\ p_{\sigma(n,q_n)}\right)\)
and where for all \(i~\in~\enum1n\) we have \(q_i \in \mathbb{N}\) and the
arguments of \(m_i \in \M{}\) consist in $q_i$ of the previously quantified
mixin parameters $p_k$, i.e. for all \(k \in \enum1{q_i}\), we
have \(\sigma(i,k) \in \enum1{i-1}\).
\end{definition}

@@

\begin{remark}For all \(i \in \enum1n\), we have
  \(\dep{}(m_i) = \enum{m_{\sigma(i,1)}}{m_{\sigma(i,q_i)}}\).
\end{remark}

@@

\begin{proposition}[\dep{} is transitively closed and describes a DAG]
  \[\forall m \in \M{}, \qquad \dep{}^\star(\dep{}^\star(m)) \subseteq
  \dep{}^\star(m) \quad \textrm{and} \quad m \notin \dep{}(m).\]
\end{proposition}
\begin{proof}
  Indeed \dep{} is transitively closed because records are well typed
  in the empty context and describes a DAG since Coq does not admit
  circular definitions.
\end{proof}

\begin{definition}[\factories{} \F{}]
A \emph{\factory{}} \(f \in \F{}\) is a distinguished Coq record or definition with one or more parameters.
The first parameter must be a \coq{(T : Type)}, while the other parameters
are $n$ \mixins{}, applied to \coq{T} and previous mixin variables. I.e.
\begin{coqcode}
Record (* or Definition *) #$f$# (T : Type) #\(\pmp{n}\)# : Type := ...
\end{coqcode}
\end{definition}

\begin{definition}[$\requires{}$, factory requirements.]
  Given \(f \in \F{}\), we define \(\requires{}(f) \in \powerset\M{}\)
  as the set of all \mixins{} that occur as parameters of \(f\), i.e.
  \(\dep{}(f) = \enum{m_1}{m_n}\).
\end{definition}

The following property holds,
\begin{proposition}[\requires{} is closed under \dep{}]
\(\forall f \in \F{}, \dep{}^\star(\requires{}(f)) \subseteq \requires{}(f)\)
\end{proposition}
\begin{proof}
Because Coq records and global definitions are well typed in the empty context.
\end{proof}

\begin{definition}[\mixinbuilders{} \B{}]
  A \emph{\mixinbuilder{}} \(\mu \in \B{}\) is a distinguished function whose
  return type is a mixin \(m_{n+1} \in \M{}\) and whose parameters are
  the carrier type \coq{(T : Type)}, \mixins{}
  \(\enum{m_1}{m_n} \in \powerset\M{}\) and a \emph{\factory{}}
  \(f \in \F{}\), such that \( \requires{}(f) =
  \enum{m_1}{m_n}\). I.e.
\begin{coqcode}
Definition #$\mu$# (T : Type) #$\pmp{n}$#: #\(f\ T\ p_1 \ldots p_n\)# -> #\(m_{n+1}\ T\ p_{\sigma(n+1,1)}\ \ldots\ p_{\sigma(n+1,q_{n+1})}\)# := ...
\end{coqcode}
\end{definition}

Note that the \mixinbuilders{} of a given \factory{} have the same set
of dependencies.

\begin{definition}[\from]
  We define \(\from{} (f, m_{n+1}) = \mu\) to be the (unique) \mixinbuilder{}
  for \(m_{n+1}\) from the \factory{} \(f\), when it exists.
\end{definition}

Note that \(\from\) is not a total function, and that \(\from{}(f, m)\) is
defined if and only if there is a declared builder that shows how to
build $m$ from $f$. In this case, we say $f$ provides $m$.

\begin{definition}[$\provides{}$]
  \( \provides{}(f) = \{m | \from{}(f,m) \textrm{ is defined}\}
    \in \powerset\M{}\) the set of mixins that a factory
    \(f \in \F{}\) provides, through its \mixinbuilders{}.
\end{definition}

@@

\begin{proposition}[\( \M{} \subseteq \F{} \)] There is a way to see mixins as factories.
\end{proposition}
\begin{proof}
  For all \(m \in \M{}\) we have \(\requires{}(m) = \dep{}(m)\),
  \(\provides{}(m) = \{m\}\) and
  \(\from{}(m,m) = \left(\coq{fun T }\ \pmp{n} \ \coq{(x :}\ m\
    \coq{T }\ p_1\ \ldots p_n\ \coq{) => x}\right) \in \B{}.\)
\end{proof}

\begin{command} to declare a new mixin:
\begin{coqcode}
#\hbmixin{}# Record M T of #$f_1\ \ldots\ f_n$# := { .. }.
\end{coqcode}
  Creates a module \coq{M} with a record which is exported as
  \coq{M}, and which depends on the \mixins{}
  \(\requires{}(\coq{M}) = \dep{}(\coq{M}) =
  \provides{}^\star\left(\enum{f_1}{f_n}\right)\), and registers this
  new record both as a \mixin{} and a \factory{}. It also exports an
  \phantterm{} \coq{M.Build.}
\end{command}

\begin{command} to declare an instance: \hbinstance{}
  \(X\ b_1 \ldots b_k\), synthesizes terms corresponding to all the
  \mixins{} that can be built from the $b_i$. Indeed if
  \(b_i : f_i\ T\ \ldots\), then this command creates elements of types
  \(\provides{}^\star\left(\enum{f_1}{f_k}\right).\) This command also
  generates unification hints as described in
  Section~\ref{sec:target-lang}.
\end{command}

\begin{commands} to declare a new \factory{} and generate new \mixinbuilders:
\begin{coqcode}
#\hbfactory{}# Record F T of #$f_1\ \ldots\ f_n$# := { .. }.
#\hbbuilders{}# Context T (a : F T).

  Definition #$b_{n+1}$# : #$f_{n+1}$# T .. := ...
  #\hbinstance{}# T #$b_{n+1}$#.
  ..
  Definition #$b_{n+k}$# : #$f_{n+k}$# T .. := ...
  #\hbinstance{}# T #$b_{n+k}$#.
#\hbend{}#.
\end{coqcode}
  Creates a module \coq{F} with a record which is exported as
  \coq{F}, and which depends on the mixins
  \(\requires{}(\coq{M}) =
  \provides{}^\star\left(\enum{f_1}{f_n}\right)\) and registers this
  new record as a factory.  It also exports an
  \phantterm{} \coq{F.Build.} We then use the \factoryinstances{}
  \(b_{n+1}, \ldots, b_{n+k}\) provided by the user and derive
  \mixinbuilders{} from them, so that
  \(\provides{}(\coq{F}) = \provides{}^\star\left(\enum{f_{n+1}}{f_{n+k}}\right).\)

  It is thus necessary that
  \(\requires{}^\star\left(\enum{f_{n+1}}{f_{n+k}}\right) \subseteq
  \provides{}^\star\left(\enum{f_1}{f_n}\right).\)
\end{commands}

Note that the \(b_i\) are not \mixinbuilders{} since their return types
are not necessarily \mixins{}, but could be \factories{}. However, since all
\factories{} provide \mixins{} through \mixinbuilders{}, we
obtain \mixinbuilders{} out of each \(b_i\) by function composition.

@@

\begin{definition}[\C{}, class]
A \class{} \coq{c} \(\in \C{}\) is distinguished a Coq record with one parameter
\coq{(T : Type)}. The type of each field is a \mixin{} in \M{} applied to
\coq{T} and, if needed, any number of other fields:
\begin{coqcode}
Record #$c$# (T : Type) := { #\(p_1 : m_1\ T\)#; .. ; #\(p_i : m_i\ T\ p_{\sigma(i,1)}\ \ldots p_{\sigma(i,q_i)}\)#; ..}.
\end{coqcode}
\end{definition}

@@

\begin{definition}[\Str{} Structure]
A \structure{} \coq{s} \(\in \Str{}\) is a distinguished dependent pair: a Coq record where
the value of the first field occurs in the type of the second.
The first field is \coq{(sort : Type)} and the second field is
\coq{(class : c sort)} for some \coq{c} \(\in \C{}\).
As a consequence \structures{} are in bijection with \classes{}.
\end{definition}

\begin{command} to declare a class and structure:
  \hbstructure{} \coq#Definition M := { A of# \(f_1 \ldots f_n\) \coq#}#
  crafts a class \coq{M.class_of} $\in \C{}$ where
  $\cdef{}(c) = \provides{}^\star\left(\enum{f_1}{f_n}\right) $ and the
  corresponding structure \coq{M.type} $\in \Str{}$, together with
  unification hints as described in Section~\ref{sec:target-lang}.
\end{command}
